Alex's Notes and Text Fragments for the JSS Manuscript

### 2021-09-24 paper/paper.Rmd
revised (partly rewrote) spatiotemp methods section

## Cross-Validation for spatiotemporal data

Some methods have the ability to operate in multiple spaces, i.e. spatial, temporal or spatiotemporal (2D + time).
To avoid redundancy and putting too much weight on certain methods, only the spatiotemporal examples of these methods will be shown in the following sections.
For the respective other spaces, usually only either the space or time coordinates need to be omitted from the user input.

### Custom: "Leave-location-and-time-out" --- `"spcv_cstf"`

Leave-location-and-time-out (LLTO) from package \pkg{CAST} is an implementation of spatiotemporal resampling method [@meyer2018].
It requires both a spatial and a temporal variable which are used to separate observations in a multi-dimensional space.
It comes with satellite methods named "leave-time-out" (LTO) and "leave-location-out" (LLO) which only operate in a two-dimensional space.

In this example the `cookfarm` dataset is used because it contains the variable ("Date") which serves as the temporal component in the dataset.

The sub-methods "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO) are conceptually similar to the "cv" method with a custom grouping from \pkg{mlr3} as they perform a CV at the block level using a predefined grouping structure.

@@

In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` which is an acronym originating from the underlying upstream implementation in package \pkg{CAST} `CreateSpacetimeFolds()`.

#### Leave-Time-Out (LTO)

To better showcase the partitioning of the "Leave-Time-Out" (LTO) method, the temporal variable `Date` was adjusted in this example.

The following code uses the `cookfarm` dataset as the base, modifies the `Date` variable in such a way that it contains five unique levels and then creates a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:lto)).

@@

```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-time-out CV from package CAST (method "sptcv\\_cstf" and argument time\\_var = "Date"). Only five folds and five time points were used in this example. Note that the blue dots correspond to five discrete time levels, which appear as a point cloud due to the viewing angle.', fig.pos="ht"}
knitr::include_graphics("png/lto.png")
```

#### Leave-Location-Out (LLO)

Method "Leave-Location-Out" (LLO) works similar to "Leave-Time-Out" (LTO) in that it operates in a two-dimensional space.
Here, specific locations are left out individually for forming test partitions while the temporal information is ignored (Figure&nbsp;\@ref(fig:llo)).

@@

```{r llo, echo=FALSE, out.width="70%", fig.cap='Perspective plot of leave-location-out CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID"). The data is viewed from the future towards the past.', fig.pos="ht"}
knitr::include_graphics("png/llo.png")
```

#### Leave-Location-and-Time-Out (LLTO)

The "Leave-Location-and-Time-Out" (LLTO) method combines both "Leave-Time-Out" (LTO) and "Leave-Location-Out" (LLO).
It creates partitions that operate on spatiotemporal clusters to reduce both spatial and temporal autocorrelation.
If there are multiple observations over time per location, and the modeling purpose is to reduce the bias in the predictions as potentially introduced by spatial as well as temporal autocorrelation, LLTO might be an appropriate choice.
To reduce autocorrelation in both dimensions, the methods omits observations as the splitting in test and train sets only provides separation in one dimension whereas two (space and time) are needed to account for both types, spatial and temporal autocorrelation.
In other words, accounting for one dimension as in Figure&nbsp;\@ref(fig:lto) or Figure&nbsp;\@ref(fig:llo), by only having one time step or location in either test or training set, would have no effect on the respective other.
Thus, the removal of observation can be seen as a proxied buffering in multiple dimensions as shown in Figure&nbsp;\@ref(fig:llto).

>>>>

## Cross-validation for spatiotemporal data

Some of the implemented resampling methods operate in multiple dimensions, i.e. in space, time, or space--time.
In this section, only examples of these methods in the spatiotemporal domain will be shown.
For their application in lower dimensions, usually only either the space or time coordinates need to be omitted from the user input.

### Custom: "Leave-location-and-time-out" and related methods --- `"spcv_cstf"`

@meyer2018 proposed a spatiotemporal resampling method in which a test set is selected and all observations that correspond to the same location or time point are omitted from the training sample.
This method is referred to as "leave-location-and-time-out" (LLTO) in package \pkg{CAST}.
<!-- AB: The following information is kinda obvious: -->
<!-- It requires both space coordinates and time point identifiers that are used to locate observations. --> 
Additional methods that resample in the temporal and spatial domain only are named "leave-time-out" (LTO) and "leave-location-out" (LLO), respectively, in \pkg{CAST}.
Note that despite their names, LLTO, LTO and LLO are conceptually not leave-*one*-out methods as they place a certain fraction of observations in the test set, as in ordinary CV.
Also, LTO and LLO are conceptually similar to \pkg{mlr3}'s "cv" method with a custom grouping as they perform a CV at the block level using a grouping structure defined by time points (LTO) and locations (i.e., time series; LLO).

In this section the `cookfarm` dataset is used as an example because has a temporal dimension identified by the variable "Date".

@@

<!-- AB: No need to justify internal names of methods:
In \pkg{mlr3spatiotempcv} the method is named `"sptcv_cstf"` which is an acronym originating from the underlying upstream implementation in package \pkg{CAST} `CreateSpacetimeFolds()`.
-->

#### CV at the time-point level: "leave-time-out" (LTO)

In the LTO method, the time points are resampled into the desired number of folds.
In the terminology used in this work, this can be referred to as resampling at the level of time points, which effectively define blocks.
Thus, observations from the same time point are jointly sampled into the same test (or training) fold, with no constraints on the temporal distance between the sampled time points.
This method does therefore not implement block CV in the sense of the time series literature.

In the `cookfarm` example, we reduce the `Date` variable to five unique levels for better visualization, and then create a spatiotemporal regression task in \pkg{mlr3spatiotempcv} (Figure&nbsp;\@ref(fig:lto)).

@@

```{r lto, echo=FALSE, out.width="70%", fig.cap='Perspective plot of "leave-time-out" CV from package CAST (method "sptcv\\_cstf" and argument time\\_var = "Date"). Only five folds and five time points were used in this example. Note that the blue dots correspond to five discrete time levels, which appear as a point cloud due to the viewing angle.', fig.pos="ht"}
knitr::include_graphics("png/lto.png")
```

#### CV at the location level: "leave-location-out" (LLO)

In contrast to LTO, the LLO method randomly resamples locations that may, for example, correspond to time series.
The sampled locations form the test partition while the temporal information is ignored (Figure&nbsp;\@ref(fig:llo)).
Unlike spatial CV methods that are based on geometric regions or the clustering of coordinates, the sampled test locations bear no particular spatial relationship.

@@

```{r llo, echo=FALSE, out.width="70%", fig.cap='Perspective plot of "leave-location-out" CV from package CAST (method "sptcv\\_cstf" and argument space\\_var = "SOURCEID"). The data is viewed from the future towards the past.', fig.pos="ht"}
knitr::include_graphics("png/llo.png")
```

#### "Leave-location-and-time-out" (LLTO)

In LLTO, a test set is first randomly sampled from the data set, and then all observations that correspond to the same location or time point are omitted from the training sample (Figure&nbsp;\@ref(fig:llto)).
LLTO resampling mimics the situation where a model is trained on time series data from a number of locations and time points, and used to predict the time series at other locations and time points that are not included in the training sample.

Conceptually, LLTO applies zero-distance buffering in both space and time: The buffer zones consist of all observations whose distance to the test sample in either space or time equals zero.
In the mathematical sense, however, this buffering is not based on a valid metric (or distance function) in three-dimensional space (2D + time) as neither the identity of discernables nor the triangle inequality are satisfied by the underlying combined 'distance' measure.
Also note that LLTO does not 'combine' LTO with LLO, as neither of these applies a buffer zone.
